\documentclass[11pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\begin{document}

	\section*{Biographical note}
	\begin{itemize}
		\item Bohdan Khomtchouk is currently an Instructor at University of Chicago, previously an American Heart Association Postdoctoral Fellow at Stanford University. His principal research focus has been on creating artificial intelligence and machine learning software to organize and better understand the world's cardiovascular disease knowledge (text and data) at a massive scale -- working at the multidisciplinary interface of healthcare, algorithm design, software engineering, integrative bioinformatics, multi-omics, natural language processing, and statistical learning. 
		
		\item Diem-Trang Tran is a doctoral student in Data Management and Analysis at the School of Computing, University of Utah, where she works on a variety of bioinformatics analyses, ranging from processing, visualizing and quantitative analyses of high-throughput sequencing data.
		
		\item Kasra Vand is a Research Scientist at Quiltomics and one of the lead developers of the HeartBioPortal project.  He holds a Bachelor's degree in Software Engineering and his research interests span broadly across AI, machine learning, and natural language processing.
		
		\item Matthew Might has been the Director of the Hugh Kaul Precision Medicine Institute at the University of Alabama at Birmingham (UAB) since 2017.  At UAB, Dr. Might is the Hugh Kaul Endowed Chair of Personalized Medicine, a Professor of Internal Medicine and a Professor of Computer Science.
		
		\item Or Gozani is the Morris Herzstein Professor of Biology at Stanford University.  The main focus of his laboratory is understanding the molecular mechanisms by which chromatin signaling networks affect nuclear and epigenetic programs, and how disruption in these mechanisms contribute to cancer and other pathologies including heart disease.
		
		\item Themistocles (Tim) Assimes is an Associate Professor at Stanford University.  He is a board certified clinical cardiologist whose lab investigates the genomic determinants of coronary heart disease (CHD) and risk factors of CHD.  Dr. Assimes is also interested in clinical prediction modeling using genetic and non-genetic biomarkers.
	\end{itemize}
\end{document}