---
title: "Pubmed"
author: "Trang Tran"
date: "8/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(rentrez)
library(GEOmetadb)
library(RSQLite)
library(magrittr)

count_year <- function(year, term) {
    query <- paste(term, "AND (", year, "[PDAT])")
    return(entrez_search(db="pubmed", term=query, retmax =0)$count)
}

get_pmids <- function(query) {
    entrez_search(db="pubmed", term=query, retmax =0)$count %>%
        entrez_search(db="pubmed", term=query, retmax = .) %>%
        `$`('ids') %>%
        return()
}

```

```{r readData}
heartConditions = read.table('../data/conditionList.tsv', sep='\t', header = TRUE)
heartPubmed = list()
endYear = format(Sys.time(), "%Y") %>% as.integer()
```

## Pubmed

### Queries

```{r, include=FALSE, eval=TRUE}
queries <- c('cardiovascular diseases' = '"cardiovascular diseases"[MeSH Terms]',
             'bioinformatics' = 'bioinformatics[MeSH Terms] OR genomics[MeSH Terms]',
             'cardioinformatics' = '(bioinformatics[MeSH Terms] OR genomics[MeSH Terms]) AND ("cardiovascular diseases"[MeSH Terms])',
             'cancer informatics' = '(bioinformatics[MeSH Terms] OR genomics[MeSH Terms]) AND (*cancer[MeSH Terms])',
             'cancer' = '(*cancer[MeSH Terms])'
              )

# Reviewer's queries
queries['bioinformatics_b'] = 'bioinformatics OR (computational biology) OR (machine learning) OR (systems biology) OR (systems medicine) OR (precision medicine)'
queries['cardioinformatics_b'] = glue::glue('({queries["cardiovascular diseases"]}) AND ({queries["bioinformatics_b"]})')
queries['cancer informatics_b'] = glue::glue('({queries["cancer"]}) AND ({queries["bioinformatics_b"]})')

for (i in 1:length(queries)) {
    queries[i] = paste('((', queries[i], 'AND (hasabstract[text] AND English[lang]))) NOT (("autobiography"[Publication Type] OR "biography"[Publication Type] OR "corrected and republished article"[Publication Type] OR "duplicate publication"[Publication Type] OR "electronic supplementary materials"[Publication Type] OR "interactive tutorial"[Publication Type] OR "interview"[Publication Type] OR "lectures"[Publication Type] OR "legal cases"[Publication Type] OR "legislation"[Publication Type] OR "meta analysis"[Publication Type] OR "news"[Publication Type] OR "newspaper article"[Publication Type] OR "patient education handout"[Publication Type] OR "published erratum"[Publication Type] OR "retracted publication"[Publication Type] OR "retraction of publication"[Publication Type] OR "review"[Publication Type] OR "scientific integrity review"[Publication Type] OR "support of research"[Publication Type] OR "video audio media"[Publication Type] OR "webcasts"[Publication Type]))'
)
}
```

Random check if query results are expected

```{r,eval=FALSE}
queryPMIDs <- lapply(queries, get_pmids)
saveRDS(queryPMIDs,file='../data/pmids_by_query.RDS')
randomSubsetQueries <- lapply(queryPMIDs, function(q) {
    sample(q,20,replace = FALSE) %>%
        sapply(function(x) {return(paste0(x,"[uid]"))}) %>%
        paste(collapse =" OR ") %>%
        return()
        })
data.frame('sector' = names(queries),
           'query' = queries,
           'random20' = randomSubsetQueries %>% as.character(),
           'random20url' = randomSubsetQueries %>% as.character() %>% URLencode() %>% paste0('https://www.ncbi.nlm.nih.gov/pubmed/?term=', .),
           stringsAsFactors = FALSE) %>%
    write.csv('../data/random20pubmed.csv')
```

### Publication on Cardiovascular diseases over the years

```{r,eval=FALSE}
relativePubmed <- data.frame('year' = c(1990:2019)) %>%
    `$<-`('all', sapply(.[['year']], function(x) { return(count_year(x, ''))})) %>%
    `$<-`('cardiovascular diseases', sapply(.[['year']], function(x) { return(count_year(x, queries['cardiovascular diseases']))})) %>%
    `$<-`('bioinformatics', sapply(.[['year']], function(x) {return(count_year(x, queries['bioinformatics']))})) %>%
    `$<-`('cardioinformatics',  sapply(.[['year']], function(x) {return(count_year(x, queries['cardioinformatics']))}))
```
```{r,eval=FALSE}
relativePubmed <- relativePubmed %>%
    `$<-`('cancer informatics',  sapply(.[['year']], function(x) {return(count_year(x, queries['cancer informatics']))})) %>%
    `$<-`('cancer',  sapply(.[['year']], function(x) {return(count_year(x, queries['cancer']))}))

```
```{r,eval=FALSE}
relativePubmed <- relativePubmed %>%
    `$<-`('bioinformatics_b',  sapply(.[['year']], function(x) {return(count_year(x, queries['bioinformatics_b']))})) %>%
    `$<-`('cancer informatics_b',  sapply(.[['year']], function(x) {return(count_year(x, queries['cancer informatics_b']))})) %>%
    `$<-`('cardioinformatics_b',  sapply(.[['year']], function(x) {return(count_year(x, queries['cardioinformatics_b']))}))
```


```{r}
sector.colors = c('cardiovascular diseases' = '#fbb4ae', 'cancer' = '#fbb4ae',
                  'bioinformatics' = '#fed9a6', 'cardioinformatics' = '#b3cde3', 'cancer informatics' = '#ccebc5',
                  'bioinformatics_b' = '#fed9a6', 'cardioinformatics_b' = '#b3cde3', 'cancer informatics_b' = '#ccebc5')
sector.orders = relativePubmed[relativePubmed$year == 2016, -1] %>% as.numeric() %>% order(decreasing = T) %>% `[`(names(relativePubmed)[-1], .)
plot_pubcount <- function(data, yearsToPlot, varsToPlot) {
    p <- data %>%
        reshape2::melt(id.vars='year') %>%
        subset(variable %in% varsToPlot & year %in% yearsToPlot) %>%
        `$<-`('variable', factor(.[['variable']], levels=sector.orders)) %>%
        plyr::arrange(variable) %>%
        ggplot() +
            geom_bar(aes(x=year,y=value,group=variable, fill=variable),stat='identity', position='identity') +
            ylab('Number of publications on Pubmed') +
            scale_x_continuous(breaks = yearsToPlot, labels=as.character(yearsToPlot)) +
            scale_fill_manual(values=sector.colors) +
            theme(axis.text.x = element_text(hjust=1, angle=45), legend.position = c(0.15, 0.85)) +
            guides(fill=guide_legend(title=""))
    if (endYear %in% yearsToPlot)
        p <- p + annotate("text", x=endYear, y=max(subset(data, year == endYear, varsToPlot))*1.1, label="*", size=7)
    return(p)    
}
```


#### Cardiovascular disease vs bioinformatics/genomics research since 2000 (when bioinformatics research starts to gather enough publication to be visible)

```{r}
yearsToPlot = 2000:2019
plot_pubcount(relativePubmed, yearsToPlot, varsToPlot = c('bioinformatics', 'cardioinformatics', 'cardiovascular diseases') )
ggsave('../figures/pubmed-cardio.png', device = 'png', width = 7, height=5)
```

#### Just curious, how is it with cancer research?


```{r}
plot_pubcount(relativePubmed, yearsToPlot = yearsToPlot, varsToPlot = c('cancer', 'bioinformatics', 'cancer informatics') ) +
    guides(fill=guide_legend(title=""))
ggsave('../figures/pubmed-cancer.png', device = 'png', width = 7, height=5)
```

#### The laggging of cardiovascular research in adopting bioinformatics/genomics technique

```{r}
plot_pubcount(relativePubmed, yearsToPlot = yearsToPlot, varsToPlot = c('bioinformatics', 'cancer informatics', 'cardioinformatics') )
ggsave('../figures/pubmed-cardioinfo-vs-cancerinfo.png', device = 'png', width = 7, height=5)
```

#### Query result with Reviewer 1's terms

```{r}

plot_pubcount(relativePubmed, yearsToPlot = yearsToPlot, varsToPlot = c('bioinformatics_b', 'cancer informatics_b', 'cardioinformatics_b') )

```

### Standard terms

```{r}
heartConditions$PubmedCount = 0
heartConditions$PubmedCount.MeSH = 0
for (i in 1:nrow(heartConditions)) {
    # tmp <- entrez_search(db='pubmed', term=paste0("(", heartConditions[i,'HPOTermName'], ")[MeSH]"))
    # heartConditions[i, 'PubmedCount.MeSH'] <- tmp$count
    heartPubmed[[i]] <- entrez_search(db='pubmed', term=paste0('("', heartConditions[i,'HPOTermName'], '")'))
}
```

```{r}
heartPubmed.Year <- list()
for (i in 1:nrow(heartConditions)) {
    cnts = sapply(c(1990:2017), function(y) { return(count_year(y, heartConditions[i, 'HPOTermName']))})
    heartPubmed.Year[[i]] = data.frame(year=c(1990:2017),
                                       count=cnts)
}
```

```{r}
for (i in 1:length(heartPubmed.Year)) {
    heartPubmed.Year[[i]]$HPOTermName = heartConditions[i,'HPOTermName']
}
p <- do.call('rbind',heartPubmed.Year) %>%
    ggplot() +
    geom_area(aes(x=year, y=count, group=HPOTermName,fill=HPOTermName,postion='stack')) +
    theme(legend.position = 'bottom')

p
plotly::ggplotly(p)
    
```

```{r}
ggplot(heartConditions) +
    geom_bar(aes(x=HPOTermName,y=PubmedCount), stat='identity', alpha=0.5) +
    geom_bar(aes(x=HPOTermName,y=PubmedCount.MeSH), stat='identity', alpha=0.5, color=2) +
    theme(axis.text.x = element_text(angle=30,hjust=1)) +
    scale_y_log10()
```

```{r}
lapply(heartPubmed, function(x) { return(x$count)})
```
```{r}
p <- ggplot(heartConditions) +
    geom_point(aes(x=PubmedCount, y=PubmedCount.MeSH)) 
ggplotly(p)
```

### Common names

```{r}
heartRegularPubmed <- 
    heartConditions$query %>%
    unique() %>%
    lapply( function(x) {
        return(entrez_search(db='pubmed', term=paste0('("', x, '")')))  
    })
```

```{r}
heartConditions$query %>%
    unique() %>%
    data.frame() %>%
    `$<-`('PubmedCount', sapply(heartRegularPubmed, function(x) {return(x$count)})) %>%
    `names<-`(c('Query', 'PubmedCount')) %>%
    ggplot() +
        geom_bar(aes(x=Query,y=PubmedCount), stat='identity') +
        theme(axis.text.x = element_text(angle=45,hjust=1))
```

