---
title: "CVD GEO data sets"
author: "Trang Tran"
date: "9/5/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(rentrez)
library(GEOmetadb)
library(RSQLite)
library(magrittr)
library(dplyr)
library(tidyr)

count_year <- function(year, term) {
    query <- paste(term, "AND (", year, "[PDAT])")
    return(entrez_search(db="pubmed", term=query, retmax =0)$count)
}

get_pmids <- function(query) {
    entrez_search(db="pubmed", term=query, retmax =0)$count %>%
        entrez_search(db="pubmed", term=query, retmax = .) %>%
        `$`('ids') %>%
        return()
}

get_year <- function(submission_date) {
    return(as.numeric(gsub('^(\\d{4})(.+)$','\\1', submission_date)))
}

```

## GEO

### Datasets deposited by CVD publications

```{r}
cardioPubmed <-
    entrez_search(db='pubmed', term=queries['cardiovascular diseases'], retmax=0)$count %>%
    entrez_search(db='pubmed', term=queries['cardiovascular diseases'], retmax=.)
```

GDS is a curated version of GSE, yet becoming deprecated (Sean Davis (NIH)'s answer on Biostar). So we'll stick to GSE.

```{r}
if(!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
con <- dbConnect(SQLite(), 'GEOmetadb.sqlite')

cardio_gsm <- dbGetQuery(con, paste('SELECT gsm.gsm, gsm.source_name_ch1, gsm.organism_ch1, organism_ch2, gpl.gpl, gpl.title as platform, gpl.description as platform_description, technology, gse.gse, gse.submission_date, gse.pubmed_id, gse.title, gse.summary, gse.type as study_type, gsm.type as sample_type, gsm.molecule_ch1',
                      'FROM gse ',
                      'JOIN gse_gsm ON gse.gse = gse_gsm.gse',
                      'JOIN gse_gpl ON gse.gse = gse_gpl.gse',
                      'JOIN gsm ON gse_gsm.gsm = gsm.gsm',
                      'JOIN gpl ON gse_gpl.gpl = gpl.gpl',
                      'WHERE pubmed_id IN (', paste(cardioPubmed$ids,collapse=','), ')'))
```

#### Study types

```{r}
types = cardio_gsm$type %>% 
    strsplit(split = ";\t") %>%
    do.call('c',.) %>%
    unique()
technologies = cardio_gsm$technology %>% unique()
cardio_gsm$year <- as.numeric(gsub('^(\\d{4})(.+)$','\\1', cardio_gsm$submission_date))

```

#### Technology

Distribution of the number of rows in a platform by technology

```{r}
all_gpl = dbGetQuery(con, paste('SELECT gpl, technology, data_row_count, organism, title, status, submission_date',
                                'FROM gpl'))
all_gpl$submission_date <- as.Date(all_gpl$submission_date)
ggplot(all_gpl) +
    geom_boxplot(aes(x=technology,y=data_row_count,group=technology)) +
    # geom_point(aes(x=technology,y=data_row_count, color = submission_date), position = 'jitter', alpha=0.3) +
    scale_y_log10() +
    theme(axis.text.x = element_text(angle=30, hjust=1))
```

Time line of submission of samples in CVD research

```{r}
ggplot(cardio_gsm) +
    facet_grid(technology ~ .) +
    geom_histogram(aes(x=as.Date(submission_date)))
 
```

Technology over the years, count by number of samples

```{r}
tech_gsm_counts <- cardio_gsm %>%
    subset(TRUE,c('year', 'technology', 'gsm')) %>%
    group_by(year, technology) %>%
    summarize(count = length(year)) %>%
    spread(technology, count) %>%
    reshape2::melt(id.vars='year') %>%
    set_names(c('year', 'technology', 'count'))
   
tech_gsm_counts$count[is.na(tech_gsm_counts$count)] = 0
ggplot(tech_gsm_counts) +
    # geom_line(aes(x=year,group=technology,color=technology, y=count),alpha=0.5)
    geom_area(aes(x=year,y=count,fill=technology), alpha=0.5, position='stack')

```

Cummulative number of samples

```{r}
tech_gsm_counts %>% group_by(technology) %>% arrange(year) %>% mutate(cumcount = cumsum(count)) %>%
    ggplot() +
    geom_area(aes(x=year,y=cumcount,fill=technology), alpha=0.9, position='stack') +
    ylab('Cumulative number of samples')
ggsave(filename = '../figures/gsm_count_by_tech.png', device='png', width=7, height=3)
```

Technology over the years, count by number of data sets

```{r}
tech_gse_counts <- cardio_gsm %>%
    subset(TRUE,c('year', 'technology', 'gse')) %>% unique() %>%
    group_by(year, technology) %>%
    summarize(count = length(year)) %>%
    spread(technology, count) %>%
    reshape2::melt(id.vars='year') %>%
    set_names(c('year', 'technology', 'count'))
tech_gse_counts$count[is.na(tech_gse_counts$count)] = 0
ggplot(tech_gse_counts) +
    # geom_line(aes(x=year,group=technology,color=technology, y=count),alpha=0.5)
    # geom_bar(aes(x=year,y=count,fill=technology), stat='identity',alpha=0.9, position='stack', width=1)
    geom_area(aes(x=year,y=count,fill=technology), stat='identity',alpha=0.9, position='stack') +
    ylab('Number of data sets') +
    ggsave(filename = '../figures/gse_count_by_tech.png', device='png', width=7, height=3)
```



GEO Data from human samples over the year

```{r}
cardio_gsm_human <- cardio_gsm %>%
    `$<-`('year', as.numeric(gsub('^(\\d{4})(.+)$','\\1', .$submission_date))) %>%
    subset(grepl('Homo sapiens', .$organism_ch1))
data.table::fwrite(cardio_gsm_human, file='../data/cardio_gsm_human.tsv',sep='\t')
cardio_gsm_human %>%
    subset(TRUE,c('year', 'organism_ch1', 'gsm')) %>%
    ggplot() +
        geom_line(aes(x=year,group=organism_ch1,color=organism_ch1), stat='count',alpha=0.5)
```

```{r}

cardiovascular_gse <- dbGetQuery(con,paste("select gse,type,pubmed_id,summary,title from gse where",
                                           "summary like '%cardiovascular%'", sep=" "))
cardiovascular_gds <- dbGetQuery(con,paste("select gds,pubmed_id,gse,platform_organism,sample_organism,description,title from gds where",
                                            "description like '%cardiovascular%'", sep=" "))
str(cardiovascular_gds)
```

### Data sets from studies not originally targeted for CVD

Samples from studies not specifically targeted to cardiovascular diseases can be re-purposed for CVD research, for example tissue samples from heart, vein, blood vessels, etc.

```{r}
all_gsm = dbGetQuery(con, paste('SELECT gsm, gsm.source_name_ch1, gsm.source_name_ch2, gsm.type as sample_type, gsm.organism_ch1, gsm.organism_ch2, gsm.title, gsm.submission_date, gsm.status, gpl.gpl, gpl.title as platform, gpl.technology, gsm.molecule_ch1',
                                'FROM gsm',
                                'INNER JOIN gpl ON gsm.gpl = gpl.gpl'))
all_gsm$year <- get_year(all_gsm$submission_date)
```

```{r}
heart_related_idx <- ((grepl(pattern = '[Hh]eart', x = all_gsm$source_name_ch1) |
                          grepl(pattern = '[Aa]rtery', x = all_gsm$source_name_ch1)) &
                          grepl(pattern = 'Homo sapiens', x = all_gsm$organism_ch1)) %>%
    which()
```

The number of samples in non-CVD studies but can be of interest `r all_gsm[heart_related_idx, 'gsm'] %>% setdiff(cardio_gsm$gsm) %>% length()`

```{r}
all_gsm[heart_related_idx,] %>% `$`('sample_type') %>% table()
```

```{r}
all_gsm[heart_related_idx, ] %>% `$`('molecule_ch1') %>% table() %>% data.frame() %>% set_names(c('molecule', 'non-CVD studies')) %>%
    merge(cardio_gsm %>% `$`('molecule_ch1') %>% table() %>% data.frame() %>% set_names(c('molecule', 'CVD studies')), by = 'molecule', all = TRUE)
```

```{r}
cardio_gsm %>% `$`('sample_type') %>% table()
```

```{r}
all_gsm %>%
    `[`(heart_related_idx,TRUE) %>% group_by(technology) %>% arrange(year) %>% mutate(cumcount = cumsum(count))
    ggplot() +
    geom_area(aes(x=year,y=cumcount,fill=technology), alpha=0.9, position='stack') +
    ylab('Cumulative number of samples')
ggsave(filename = '../figures/gsm_count_by_tech.png', device='png', width=7, height=3)
```

## dbGaP

```{r}
dbgap.gru = data.table::fread('../data/dbgap-gru-datasets.tsv', sep='\t')
subject.cnt.cvd = dbgap.gru[dbgap.gru$Topic == 'CVD',c(-1,-2, -3)] %>%
    summarise_all(funs(sum))
subject.cnt.all = dbgap.gru[,c(-1,-2, -3)] %>%
    summarise_all(funs(sum))
rbind(subject.cnt.cvd, subject.cnt.all) %>%
    t() %>%
    set_colnames(c('CVD', 'All'))
    reshape2::melt() %>%
    ggplot(aes(x = variable, y = value)) +
    geom_bar(stat = 'identity') +
    coord_flip()
ggplot(dbgap.gru, mapping = aes())
```

```{r}
dbgap = data.table::fread('../data/dbgap-adv-datasets.tsv', sep='\t')
dbgap$`Subject count` <-
    gsub(pattern = '^(.*\\W+)(\\d+)(\\W+subjects.*)$', replacement = '\\2', x = dbgap$`Study Content`) %>% as.numeric()
```

The number of subjects in all CVD studies deposited on dbGaP are `r dbgap[['Subject count']] %>% sum(na.rm = TRUE)`